Gaia's constraints on the extremely massive ($\sim$$150\,M_\odot$) runaway star candidate VFTS682

### 2018-06-20 VFTS682_APJL.tex
minor
significant minors

       \hline
       \multicolumn{5}{l}{\emph{Selected stellar parameters VFTS 682}}
      \\
      \hline
     & present day mass  & $[M_\odot]$ & $137.8^{+27.5}_
                                           {-15.9}$ & (1)
                                                    \\
      & initial mass& $[M_\odot]$ & $150.0^{+28.7}_{-17.4}$ & (1)
      \\
      &age & [Myr] & $1.0\pm0.2$ & (1) \\
      &mass loss rate & [$M_\odot \ \mathrm{yr}^{-1}$] & $10^{-4.1\pm0.2}$ & (2)\\
      &\multirow{2}{*}{surface helium mass fraction} &  & 0.45 & (2)\\
      & &  & 0.49 & (3)\\
      \hline

    \end{tabular}
    \tablecomments
    { 
      (1)~\cite{schneider:18}
      (2)~\cite{bestenlehner:11}
      (3)~\cite{rubio-diez:17}
    }
  \end{center}
  \label{tab:star_param}
\end{table}

@@

      \multicolumn{5}{l}{\emph{Gaia relative proper motion for VFTS682 }} \\
      \hline
      &$\delta\mu_\mathrm{RA}$  &[$\masyr$] & $0.103\pm0.080$ & (1,5) \\
      &$\delta\mu_\mathrm{DEC}$  &[$\masyr$] & $0.086\pm0.100$ &  (1,5) \\
      &$\delta\mu_{Gaia}$  &[$\masyr$] & $0.13\pm0.09$ &  (1,5) \\
      &$v_\mathrm{2d}$  &[$\kms$] & $32\pm21$ & (1,5)\\  
      &$\theta$  &[degrees] &  $14.2_{-43}^{+44}$  & (1,5)\\  
      &$v_\mathrm{2d, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
 %

 \hline     
      \multicolumn{5}{l}{\emph{HST relative proper motion for VFTS682}} \\
            \hline
      &$\delta\mu_\mathrm{RA, HST}$  &[$\masyr$] & $0.01\pm0.13$ & (4) \\
      &$\delta\mu_\mathrm{DEC, HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
       &$\delta\mu_\mathrm{HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
                  &$v_\mathrm{2d, HST}$  &[$\kms$] & $47\pm24$ & (4)\\  
                      &$\theta^{*}$  &[degrees] &   $-35_{-51}^{+24}$   & (1,5)\\  
                          &$v_\mathrm{2d, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
 %

@@

%of $137.8^{+27.5}_ {-15.9}\Msun$, an apparent age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial mass of $150.0^{+28.7}_{-17.4}\Msun$.  Further parameters of are summarized in Table~\ref{X}.

@@

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}
  \includegraphics[width=0.35\textwidth]{figures/angle}
  \caption{\emph{Left panel}: distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136. VFTS682 is not an outlier, but
    its relative proper motion matches the expected value if it were indeed
    ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. The top axis shows the conversion to physical units
    assuming a distance of 50\,kpc. \emph{Right panel}:  distribution of
    angles between the relative proper motion direction and the radial
    direction from the center of the cluster to the star. The error bars from \emph{Gaia} for VFTS682 are large, but
    the best value is in agreement with the hypothesis of dynamical
    ejection. In both
    panels, the dark blue histograms contain 317 
    stars with error smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with errors smaller than $0.05\,\mathrm{mas \
      yr^{-1}}$.\todo{possibly: cartoon to illustrate angle definition, or add to
    fig.2 if not too messy, maybe also show HST measurements on this
    plot for 682?}}
  \label{fig:dist}
\end{figure*}

@@

With these we compute the relative proper motion
$\delta\mu_\mathrm{RA}$ and $\delta\mu_\mathrm{DEC}$.  We also derive
the angle $\theta$ between the direction of motion and the vector
connecting the center of R136 with the current position of VFTS 682,
i.e. such that $\theta = 0$ is the angle corresponding to perfectly
radial motion away from R136.    We further provide the 2D velocity
projected onto the radial vector pointing away from R136
$v_\mathrm{2d, \parallel}$  and the total projected 2d velocity
$v_\mathrm{2d}$.   All kinematic quantities are provided in
Tab.~\ref{tab:vfts682}. %%what does the projection on the proper
                        %%motion in the radial direction tell us?
                        %%Seems like a useless number to me...

@@

The right panel shows the distributions in flight direction for these
stars: $\theta$ is the angle between the radial direction from the
center of R136 to each star and its proper motion relative to
R136 (see also \Figref{fig:main}). Dynamical ejections from the
cluster should %are expected to
produce
close to radial ejections, i.e.~$\theta\simeq0$. On this panel, the green band highlights
the range of angles corresponding to an opening angle around the
radial direction of 45\,degrees. The direction of motion of VFTS682 is
poorly constrained because of the large errors on the proper motion
components both in HST and \emph{Gaia} data, but
it is consistent with dynamical ejection from R136.

@@

Based on our results, we tentatively claim that VFTS682 is the most massive
(slow) runaway known to date, with a peculiar two-dimensional
projected velocity with respect to R136 of
$\sim38\pm22\kms$ (averaging the \emph{Gaia} DR2 and HST
results\todo{check numbers}). Due to the large error bars, this result will need
to be revisited with future astrometric data. % to be revisited as updated
% astrometric parameters from future \emph{Gaia} data releases become
% available.
If confirmed, it means that isolated star formation is
\emph{not} required to explain the isolation of VFTS682. Its proper motion suggests that it was ejected from the cluster R136
$0.9\pm0.6$\,Myr ago. Because of the exceptionally large mass
of this star, this raises the question of which stars must populate
the core of the cluster.

>>>>

       \hline
       \multicolumn{5}{l}{\emph{Selected stellar parameters VFTS 682}}
      \\
      \hline
     & present day mass  & $[M_\odot]$ & $137.8^{+27.5}_
                                           {-15.9}$ & (1)
                                                    \\
      & initial mass& $[M_\odot]$ & $150.0^{+28.7}_{-17.4}$ & (1)
      \\
      &age & [Myr] & $1.0\pm0.2$ & (1) \\
      &mass loss rate & [$M_\odot \ \mathrm{yr}^{-1}$] & $10^{-4.1\pm0.2}$ & (2)\\
      % &\multirow{2}{*}{surface helium mass fraction} &  & 0.45 & (2)\\
      % & &  & 0.49 & (3)\\ %% we don't use this: if we believe it's a
      % merger it's not relevant, if we don't we still don't use it
      \hline

    \end{tabular}
    \tablecomments
    { 
      (1)~\cite{schneider:18}
      (2)~\cite{bestenlehner:11}
      % (3)~\cite{rubio-diez:17}
    }
  \end{center}
  \label{tab:star_param}
\end{table}

@@

      \multicolumn{5}{l}{\emph{Gaia relative proper motion for VFTS682 }} \\
      \hline
      &$\delta\mu_\mathrm{RA}$  &[$\masyr$] & $0.103\pm0.080$ & (1,5) \\
      &$\delta\mu_\mathrm{DEC}$  &[$\masyr$] & $0.086\pm0.100$ &  (1,5) \\
      &$\delta\mu_{Gaia}$  &[$\masyr$] & $0.13\pm0.09$ &  (1,5) \\
      &$v_\mathrm{2d}$  &[$\kms$] & $32\pm21$ & (1,5)\\  
      &$\theta$  &[degrees] &  $14_{-43}^{+44}$  & (1,5)\\  
      % &$v_\mathrm{2d, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
 %

 \hline     
      \multicolumn{5}{l}{\emph{HST relative proper motion for VFTS682}} \\
            \hline
      &$\delta\mu_\mathrm{RA, HST}$  &[$\masyr$] & $0.01\pm0.13$ & (4) \\
      &$\delta\mu_\mathrm{DEC, HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
       &$\delta\mu_\mathrm{HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
                  &$v_\mathrm{2d, HST}$  &[$\kms$] & $47\pm24$ & (4)\\  
                      &$\theta^{*}$  &[degrees] &   $-35_{-51}^{+24}$   & (1,5)\\  
                          % &$v_\mathrm{2d, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
 %

@@

%of $137.8^{+27.5}_ {-15.9}\Msun$, an apparent age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial mass of $150.0^{+28.7}_{-17.4}\Msun$.  Further parameters of are summarized in Table~\ref{X}.
VFTS682 is relatively isolated in \emph{Spitzer} images, except for the extended
source S10 interpreted as star formation triggered by the wind of
VFTS682 itself by \cite{walborn:13}. In X-rays
\todo{questions: does 682 show excess X-rays (wind collisions?)}

@@

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}
  \includegraphics[width=0.35\textwidth]{figures/angle}
  \caption{\emph{Left panel}: distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136. VFTS682 is not an outlier, but
    its relative proper motion matches the expected value if it were indeed
    ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. The top axis shows the conversion to physical units
    assuming a distance of 50\,kpc. \emph{Right panel}:  distribution of
    angles between the relative proper motion direction and the radial
    direction from the center of the cluster to the star. The error bars from \emph{Gaia} for VFTS682 are large, but
    the best value is in agreement with the hypothesis of dynamical
    ejection. In both
    panels, the dark blue histograms contain 317 
    stars with error smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with errors smaller than $0.05\,\mathrm{mas \
      yr^{-1}}$.\todo{possibly: cartoon to illustrate angle definition, or add to
    fig.2 if not too messy, maybe also show HST measurements on this
    plot for 682? Redo right panel taking care of projection of the
    radial directions!It's misleading!}}
  \label{fig:dist}
\end{figure*}

@@

With these we compute the relative proper motion
$\delta\mu_\mathrm{RA}$ and $\delta\mu_\mathrm{DEC}$.  We also derive
the angle $\theta$ between the direction of motion and the vector
connecting the center of R136 with the current position of VFTS 682,
i.e. such that $\theta = 0$ is the angle corresponding to perfectly
radial motion away from R136.    We further provide % the 2D velocity
% projected onto the radial vector pointing away from R136
% $v_\mathrm{2d, \parallel}$  and
the %total
projected 2d velocity
$v_\mathrm{2d}$.   All kinematic quantities are provided in
Tab.~\ref{tab:vfts682}. %%what does the projection on the proper
                        %%motion in the radial direction tell us?
                        %%Seems like a useless number to me...

@@

The right panel shows the distributions in flight direction for these
stars: $\theta$ is the angle between the projected radial direction from the
center of R136 to each star and its proper motion relative to
R136 (see also \Figref{fig:main}). Dynamical ejections from the
cluster should %are expected to
produce
close to radial ejections, i.e.~$\theta\simeq0$. On this panel, the green band highlights
the range of angles corresponding to an opening angle around the
radial direction of 45\,degrees. The direction of motion of VFTS682 is
poorly constrained because of the large errors on the proper motion
components both in HST and \emph{Gaia} data, but
it is consistent with dynamical ejection from R136. We note that
VFTS72 has a small radial velocity, while VFTS16 (and possibly VRTS682) has
large peculiar radial velocity and therefore accurate distances are
needed to contrain the flight direction in three dimensions.

@@

Based on our results, we tentatively claim that VFTS682 is the most massive
(slow) runaway known to date, with a peculiar two-dimensional
projected velocity with respect to R136 of
$\sim38\kms$ (averaging the \emph{Gaia} DR2 and HST
results) but compatible with zero within $2\sigma$. Due to the large error bars, this result will need
to be revisited with future astrometric data. % to be revisited as updated
% astrometric parameters from future \emph{Gaia} data releases become
% available.
If confirmed, it means that isolated star formation is
\emph{not} required to explain the isolation of VFTS682. Its proper motion suggests that it was ejected from the cluster R136
$0.9\pm0.6$\,Myr ago. Because of the exceptionally large mass
of this star, this raises the question of which stars must populate
the core of the cluster.
